Calculating life expectancy and exploring missingness

### 2016-11-04 Code/1_Explore_Data_v2.Rmd

Some of the rows have Population==0, and we explore that and record below. These mostly correspond to rows with Unknown age groups that we cannot use in our analysis so I remove them. This leaves 654 rows with Populations of 0.

```{r explore_dataset}
table(dat$Population==0, useNA = "always")
#55734 cells have Populations of 0.

table(dat$Count[dat$Population==0]==0, useNA = "always")
#Of these, all have Counts of 0, except for 26 with counts of NA

table(dat$Age2 == "Unknown", dat$Population == 0, useNA = "always")
#The vast majority of cells with populations of 0 have unknown age groups

#remove this data from the data we analyze:
dat.clean <- dat[dat$Age2 != "Unknown", ]

table(dat.clean$Population == 0, useNA = "always")
#654 cells have Populations of 0 now.

table(dat.clean$Count[dat.clean$Population == 0] == 0, useNA = "always")
summary(dat.clean$Count[dat.clean$Population == 0])
#628 are equal to 0 and 26 are equal to NA

@@

#now the Counts should only equal NA when the true count is between 1 and 9.

summary(dat.clean$Crude_Rate[dat.clean$Population == 0])
table(dat.clean$Crude_Rate[dat.clean$Population==0], useNA = "always")
#whenever the population is 0, the Crude_Rate is equal to NA.
#need to think: should these be equal to 0, or NA? How will it affect the calculations?

#States with at least one strata of 0
na.pops <- dat.clean %>% select(State2, Population) %>% filter(Population == 0)
unique(na.pops$State2)

@@

**Missingness of Count (of deaths)**
If there are between 1 and 9 deaths in the strata, this information is surpressed and coded as missing. Later in this document we investigate trends in supression/missingness to understand how it differs according to state, age, race, and sex.

@@

- Set all missing counts to 1, 5, and 9.
- Set all missing counts to 1 for Whites and 9 for Blacks and vice versa.

```{r fill_in_missing_counts}
##NEED TO UPDATE THIS SO THAT CAN'T IMPUTAT COUNT > POP SIZE.
##FOLLOW UPDATE WITH RE-RUNNING THE MODELS.
dat.clean <- dat.clean %>% mutate(Count_md1 = ifelse(is.na(Count)==T, 1, Count),
                                  Count_md5 = ifelse(is.na(Count)==T, 5, Count),
                                  Count_md9 = ifelse(is.na(Count)==T, 9, Count)
                                  )

dat.clean$Count_b9w1 <- dat.clean$Count
dat.clean$Count_b9w1[is.na(dat.clean$Count) & dat.clean$Race2 == "White"] <- 1
dat.clean$Count_b9w1[is.na(dat.clean$Count) & dat.clean$Race2 == "Black"] <- 9

dat.clean$Count_b1w9 <- dat.clean$Count
dat.clean$Count_b1w9[is.na(dat.clean$Count) & dat.clean$Race2 == "White"] <- 9
dat.clean$Count_b1w9[is.na(dat.clean$Count) & dat.clean$Race2 == "Black"] <- 1

#check recoding
dat.clean$Count_grouped <- cut(dat.clean$Count, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
dat.clean$Count1_grouped <- cut(dat.clean$Count_md1, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
dat.clean$Count5_grouped <- cut(dat.clean$Count_md5, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
dat.clean$Count9_grouped <- cut(dat.clean$Count_md9, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
dat.clean$Count_b9w1_grouped <- cut(dat.clean$Count_b9w1, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
dat.clean$Count_b1w9_grouped <- cut(dat.clean$Count_b1w9, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))

@@

table(dat.clean$Count_grouped, dat.clean$Count_b9w1_grouped, dat.clean$Race2, useNA = "always")
table(dat.clean$Count_grouped, dat.clean$Count_b1w9_grouped, dat.clean$Race2, useNA = "always")

dat.clean <- dat.clean[ , !names(dat.clean) %in% c("Count_grouped", "Count1_grouped", "Count5_grouped", "Count9_grouped", 
                                                   "Count_b9w1_grouped", "Count_b1w9_grouped")]
```

>>>>

```{r unknown_age_group}
#there is an unknown age group for each combo of state-year-race-sex-age:
table(dat$Age2)

#Count of death is always equal to 0 for unknown age groups
table(dat$Age, dat$Count==0)

#remove these data from the data we analyze:
dat.clean <- dat[dat$Age2 != "Unknown", ]
```

##Population counts of 0

It is important to explore how many cells have population counts of zero because these break the Bayesian crank, and also currently breaks the life expectancy function.

```{r explore_dataset}
table(dat.clean$Population == 0, useNA = "always")
#654 cells have Populations of 0.

table(dat.clean$Count[dat.clean$Population == 0] == 0, useNA = "always")
#Of these, all have Counts of 0, except for 26 with counts of NA

@@

summary(dat.clean$Crude_Rate[dat.clean$Population == 0])
#whenever the population is 0, the Crude_Rate is equal to NA.
#need to think: should these be equal to 0, or NA? How will it affect the calculations?

#States with at least one strata of 0
na.pops <- dat.clean %>% 
  select(State2, Population) %>% 
  filter(Population == 0)

unique(na.pops$State2)

@@

**Missingness of Count (of deaths)**
If there are between 1 and 9 deaths in the strata, this information is 'coarsened' and coded as NA. Later in this document we investigate trends in supression/missingness to understand how it differs according to state, age, race, and sex.

@@

- Set all missing counts to min({1, 5, and 9}, population_size).

```{r fill_in_missing_counts}
dat.clean <- dat.clean %>% mutate(Count_md1 = ifelse(is.na(Count)==T, pmin(1, Population), Count),
                                  Count_md5 = ifelse(is.na(Count)==T, pmin(5, Population), Count),
                                  Count_md9 = ifelse(is.na(Count)==T, pmin(9, Population), Count)
                                  )

#check recoding
dat.clean$Count_grouped <- cut(dat.clean$Count, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
dat.clean$Count1_grouped <- cut(dat.clean$Count_md1, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
dat.clean$Count5_grouped <- cut(dat.clean$Count_md5, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
dat.clean$Count9_grouped <- cut(dat.clean$Count_md9, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))

@@

dat.clean <- dat.clean[ , !names(dat.clean) %in% c("Count_grouped", "Count1_grouped", "Count5_grouped", "Count9_grouped")]
```
